---
title: "geo spatial stats packages summary"
date: "11/5/2019"
output: html_document
---

```{r setup, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(RandomFields)
library(gstat)
library(sp)
library(kriging)
library(SpatialExtremes)
```

This is a quick summary of the options we have for running statistical models on spatial data. These packages were selected based on suggestions from the class or by doing a google search. I've included links to more helpful tutorials and CRAN vignettes.  

## RandomFields

Simulation and analysis of random fields: Methods for the inference on and the simulation of Gaussian fields are provided, as well as methods for the simulation of extreme value random fields.

[CRAN](https://cran.r-project.org/web/packages/RandomFields/RandomFields.pdf) 

[R Documentation](https://www.rdocumentation.org/packages/RandomFields/versions/3.1.36)

Journal Article summarizing package capabilities, including example with weather data:
[Analysis, Simulation and Prediction of Multivariate
Random Fields with Package RandomFields](file:///Volumes/GoogleDrive/My%20Drive/203_sections/v63i08.pdf)

*"RandomFields offers more than 35 dimple models and more than 30 functionals operating on simple models to construct more complex ones"*

While our textbook by Wikle et al. does not use RandomFields, it does mention it: 

*"However, there are numerous other packages in R that could be used. Among these CompRandFld and RandomFields are worth noting because of the large selection of non-separable spatio-temporal covariance functions they make available to the user."*


__some functions:__

+ __RMmatrix__ multivariate covariance model depending on one multivariate covariance model; matrix multiplication

+ __RMbiwm__ Bivariate Whittle-Matern model; corresponding covariance function only depends on the distance r>=0 between two points

+ __RFfit__ recognizes all parameters that have the value NA, and estimates them through ML

+ __RFratiotest__ approx. chi2 test or Monte Carlo likelihood ratio test; only works for Gaussian random fields 

+ __RFcrossvalidate__ fits the model to the complete data set and then the value at each location is predicted by kriging given the data at all other locations 



Example with temperature and pressure shown below:

```{r cars}
data('weather') # from RandomFields package

# parsimonious model 
nug <- RMmatrix(M = matrix(nc = 2, c(NA, 0, 0, NA)), RMnugget())
pars.model <- nug + RMbiwm(nudiag = c(NA, NA), scale = NA, cdiag = c(NA, NA), rhored = NA)

# transform coords to Euclidean distances 
Dist.mat <- as.vector(RFearth2dist(weather[, 3:4]))

PT <- weather[ , 1:2]
#pars <- RFfit(pars.model, distances = Dist.mat, dim = 3, data = PT)

# okay this gave me an error even though I was copy and pasting, might come back to it but tbh don't feel like figuring it out, might come back to it
```

## gstat

[CRAN](https://cran.r-project.org/web/packages/gstat/index.html)

[interpolation tutorial](https://rspatial.org/raster/analysis/4-interpolation.html)

[tutorial, including variograms](https://cran.r-project.org/web/packages/gstat/vignettes/gstat.pdf)

This package is useful for all spatial data, but especially for our purposes it is a simple way to create variograms and perform Kriging.

[Intro tutorial](https://rpubs.com/nabilabd/118172)

This is the main package used in the textbook by Wikle et al., which can be found on [Gouchospace](http://spacetimewithr.org/). 
See Lab 4.1 Spatio-Temporal Kriging with gstat. Not going to get into much detail because we have this text available. 

__some functions:__

+ __krige__ simple, ordinary, or universal kriging, global or local, point or block 

+ __variogram__ calculate sample variogram from data 

+ __fit.variogram__ fits a model to sample variogram 

+ __vgmST__ constructs one variogram for space and another for time, or can do one jointly, checks for minimal set of parameters 

+ __fit.StVariogram__ fits spatio-temporal variogram of a given type of spatio-temporal sample variogram

+ __krigeST__ orginary global and local and trans Gaussian spatio-temporal kriging 

## SpatialExtremes

Tools for the statistical modelling of spatial extremes using max-stable processes, copula or Bayesian hierarchical models. The data should be extreme (not daily values), marginal distribution family is fixed, and the process family is fixed.

[CRAN](https://cran.r-project.org/web/packages/SpatialExtremes/SpatialExtremes.pdf)

This [link](https://sites.lsa.umich.edu/eva2015/wp-content/uploads/sites/44/2015/07/EVA2015.pdf) has nice tutorials for Spatial GEV, Bayesian modeling, and Max-stable processes, although I'm not sure how helpful it will be for us. [source.](https://www.stat.washington.edu/peter/PASI/show.pdf)

This [link](https://www.stat.colostate.edu/graybillconference2009/Workshop%20Files/show.pdf) includes a section on **Model Selection**, including performing AIC and likelihood ratio tests.

not sure we need to use this package. 

## kriging

*Simple and highly optomized ordinary kriging algorithm to plot geographical data*
[CRAN](https://cran.r-project.org/web/packages/kriging/kriging.pdf)

__some functions:__

+ __kriging__ ordinary kriging algorithm; assumes a minumum number of observations in order to fit variogram model

+ __plot.kriging__ plot distance v. semivariance with a fitted curve that indicates model used 

## CompRandFld

Used with the Random Fields - set of procedures for the analysis of random fields using likelihood and nonstandard likelihood methods. 

[CRAN](https://cran.r-project.org/web/packages/CompRandFld/CompRandFld.pdf)

not as many tutorials available for this one

nice summary article: 
[Analyzing spatio-temporal data with R:
Everything you always wanted to know – but were
afraid to ask](https://informatique-mia.inra.fr/resste/sites/informatique-mia.inra.fr.resste/files/analyzing-spatio-temporal.pdf) 
they use gstat and CompRandFld to look at air quality data 



